---
title: "Dynamic Factor Analysis"
author: "Eli Holmes"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Dynamic Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\RR}{\mathbf{R}}
\newcommand{\xx}{\mathbf{x}}

```{r setup, include = FALSE}
# knitr options
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1234)
```

The `dfaTMB()` function allows you to fit DFAs with the same form as `MARSS(x, form="dfa")`. This has a diagonal $\QQ$ with 1 on the diagonal and a stochastic $\xx_1$ with mean 0 and variance of 5 (diagonal variance-covariance matrix). There are only 3 options allowed for $\RR$:
diagonal and equal, diagonal and unequal, and unconstrained.

Example data

```{r}
library(MARSS)
data(lakeWAplankton, package = "MARSS")
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens", "Unicells", "Other.algae")
dat <- as.data.frame(lakeWAplanktonTrans) |>
  subset(Year >= 1980 & Year <= 1989) |>
  subset(select=phytoplankton) |>
  t() |>
  MARSS::zscore()
```

Fit with MARSS
```{r}
m1.em <- MARSS(dat, model=list(R='unconstrained', m=1, tinitx=1), form='dfa', z.score=FALSE, silent = TRUE)
```

Fit with TMB. Note the syntax will be updated to match MARSS().
```{r}
library(marssTMB)
m1.tmb <- dfaTMB(dat, model=list(m=1, R='unconstrained'))
```

Compare parameter estimates
```{r}
library(tidyr)
pars <- data.frame(name = c(paste0("R", rownames(coefficients(m1.em)$R)),
                            paste0("Z", rownames(coefficients(m1.em)$Z))),
                   EM = c(coefficients(m1.em)$R, coefficients(m1.em)$Z),
  TMB = c(as.vector(m1.tmb$Estimates$R[lower.tri(m1.tmb$Estimates$R,diag=TRUE)]),
  as.vector(m1.tmb$Estimates$Z)))
pars <- pars %>% tidyr::pivot_longer(2:3, names_to = "model")
```

```{r}
library(ggplot2)
dodge <- position_dodge(width=0.5)
ggplot(pars, aes(x=name, y=value, col=model)) +
  geom_point(position=dodge) +
  ggtitle("same estimates")
```


Compare two states models.
```{r}
m1.bfgs <- MARSS(dat, model=list(R='unconstrained', m=2, tinitx=1), form='dfa', z.score=FALSE, silent = TRUE,
               method="BFGS")
m1.tmb <- dfaTMB(dat, model=list(m=2, R='unconstrained'))
c(m1.bfgs$logLik, m1.tmb$logLik)
```

```{r echo=FALSE}
ZmatGen <- function(n, m) {
  tZ <- matrix(0.5, nrow = n, ncol = m)
  if (m > 1) {
    for (i in 1:(m - 1)) {
      tZ[i, (m - (m - 2) + (i - 1)):m] <- rep(0, m - 1 - (i - 1))
    }
    return(tZ)
  } else {
    return(tZ)
  }
}
pars <- data.frame(name = c(paste0("R", rownames(coefficients(m1.bfgs)$R)),
                            paste0("Z", rownames(coefficients(m1.bfgs)$Z))),
                   EM = c(coefficients(m1.bfgs)$R, coefficients(m1.bfgs)$Z),
  TMB = c(as.vector(m1.tmb$Estimates$R[lower.tri(m1.tmb$Estimates$R,diag=TRUE)]),
  as.vector(m1.tmb$Estimates$Z[ZmatGen(nrow(dat), 2)!=0])))
pars <- pars %>% tidyr::pivot_longer(2:3, names_to = "model")
dodge <- position_dodge(width=0.5)
ggplot(pars, aes(x=name, y=value, col=model)) +
  geom_point(position=dodge) +
  ggtitle("same estimates")
```

Compare two states models with R diagonal and equal
```{r}
mod.list <- list(R='diagonal and equal', m=2, tinitx=1)
m1.bfgs <- MARSS(dat, model=mod.list, form='dfa', z.score=FALSE, silent = TRUE,
               method="BFGS")
m1.tmb <- dfaTMB(dat, model=mod.list)
c(m1.bfgs$logLik, m1.tmb$logLik)
```

## Look at a bigger data set

Compare three states models with R diagonal and unequal and 12 time series. The EM algorithm and TMB algorithms struggles to converge with DFA models. This is specific to this set of time series.
```{r}
dat2 <- rbind(dat, dat+matrix(rnorm(nrow(dat)*ncol(dat),0,0.2), nrow(dat), ncol(dat)))
dat2 <- MARSS::zscore(dat2)
mod.list <- list(R='diagonal and unequal', m=3, tinitx=1)
# Control set to match setting in MARSS.tmp()
t1.bfgs <- system.time(m1.bfgs <- MARSS(dat2, 
                                        model=mod.list, form='dfa', 
                                        z.score=FALSE, silent = TRUE, method="BFGS", 
                                        control = list(reltol = 1e-12, maxit=2000)))[1]
t1.em <- system.time(m1.em <- MARSS(dat2, model=mod.list, form='dfa', z.score=FALSE, silent = TRUE))[1]
t1.tmb <- system.time(m1.tmb <- dfaTMB(dat2, model=mod.list))[1]
t1.tmb2 <- system.time(m1.tmb2 <- dfaTMB(dat2, model=mod.list, fun.opt="optim"))[1]
```

TMB + nlminb() is much faster. But at the default convergence settings, the Kalman-Filter + BFGS ran longer and got to a higher maximum likelihood. This seems to be an issue with TMB for both the optimizers. Also the speed savings seems to be
due to nlminb() not TMB per se.
```{r}
# Log-likelihood
c(em=m1.em$logLik, kfoptim=m1.bfgs$logLik, TMBnlminb=m1.tmb$logLik, TMBoptim=m1.tmb2$logLik)
# Time
c(em=t1.em, kfoptim=t1.bfgs, TMBnlminb=t1.tmb, TMBoptim=t1.tmb2)
```

It might be an odd initial condition issue as if BFGS is started with EM, it does not find the lower log-likelihood but ends up at the higher value that EM and TMB found. Note, this varies a bit. This is for `set.seed(1234)`.
```{r}
m1 <- MARSS(dat2, model=mod.list, form='dfa', z.score=FALSE, silent = TRUE, control = list(minit=1, maxit=10))

t2.bfgs <- system.time(m2.bfgs <- MARSS(dat2, model=mod.list, form='dfa', z.score=FALSE, silent = TRUE, method="BFGS", inits = m1))[1]
c(t2.bfgs, m2.bfgs$logLik)
```

```{r}
library(broom)
df <- data.frame(t=1:37, bfgs=tidy(m1.bfgs)$estimate, em=tidy(m1.em)$estimate) |>
  pivot_longer(2:3)
ggplot(df, aes(x=t, y=value, col=name)) + geom_point()
```