---
title: "Dynamic Factor Analysis"
author: "Eli Holmes"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Dynamic Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\RR}{\mathbf{R}}
\newcommand{\xx}{\mathbf{x}}

```{r setup, include = FALSE}
# knitr options
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1234)
```

The `MARSS_tmb()` function allows you to fit DFAs with the same form as `MARSS(x, form="dfa")`. This has a diagonal $\QQ$ with 1 on the diagonal and a stochastic $\xx_1$ with mean 0 and variance of 5 (diagonal variance-covariance matrix). There are only 3 options allowed for $\RR$:
diagonal and equal, diagonal and unequal, and unconstrained.

Example data

```{r}
library(MARSS)
data(lakeWAplankton, package = "MARSS")
phytoplankton <- c("Cryptomonas", "Diatoms", "Greens", "Unicells", "Other.algae")
dat <- as.data.frame(lakeWAplanktonTrans) |>
  subset(Year >= 1980 & Year <= 1989) |>
  subset(select=phytoplankton) |>
  t() |>
  MARSS::zscore()
```

Fit with MARSS
```{r}
system.time(m1.em <- MARSS(dat, model=list(R='unconstrained', m=1, tinitx=1), form='dfa', z.score=FALSE, silent = TRUE))
```

Fit with TMB. Note the syntax will be updated to match MARSS().
```{r}
library(marssTMB)
system.time(m1.tmb1 <- dfaTMB(dat, model=list(m=1, R='unconstrained')))
```

```{r}
system.time(m1.tmb2 <- MARSS_tmb(dat, model=list(m=1, R='unconstrained')))
```

```{r}
Rprof("tmb.Rprof")
m1.tmb2 <- MARSS_tmb(dat, model=list(m=1, R='unconstrained'))
Rprof(NULL)
```
Total time `r summaryRprof("tmb.Rprof")$sampling.time` and breakdown
```{r}
summaryRprof("tmb.Rprof")$by.self
```

```{r}
Rprof("tmb.Rprof")
m1.tmb1 <- dfaTMB(dat, model=list(m=1, R='unconstrained'))
Rprof(NULL)
summaryRprof("tmb.Rprof")$by.self
```


Compare parameter estimates
```{r}
library(tidyr)
pars <- data.frame(name = c(paste0("R", rownames(coefficients(m1.em)$R)),
                            paste0("Z", rownames(coefficients(m1.em)$Z))),
                   EM = c(coefficients(m1.em)$R, coefficients(m1.em)$Z),
  TMB1 = c(as.vector(m1.tmb1$Estimates$R[lower.tri(m1.tmb1$Estimates$R,diag=TRUE)]),
  as.vector(m1.tmb1$Estimates$Z)))
pars <- pars %>% tidyr::pivot_longer(2:3, names_to = "model")
```

```{r}
library(ggplot2)
dodge <- position_dodge(width=0.5)
ggplot(pars, aes(x=name, y=value, col=model)) +
  geom_point(position=dodge) +
  ggtitle("same estimates")
```
